Supplementary Materials for CBEA: Competitive Balances for Taxonomic Enrichment Analysis

### 2022-02-01 cbea_revised.tex

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing. However, microbiome taxonomic data is sparse and high dimensional which prevents the application of standard statistical techniques. One approach to address this problem is to perform analyses at the level of taxon sets. Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability. Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given that compared to other 'omics data types microbiome taxonomic data is compositional. We developed a new tool to generate taxon set enrichment scores at the sample level through a novel log-ratio formulation based on the competitive null hypothesis. Our scores can be used for statistical inference, and as inputs to other downstream analyses such as prediction models. We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.

@@

In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate taxa to higher Linnean classification levels such as genus, family, or phylum. However, despite this interest, there is limited research done to extend existing set-based methods to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used and p-value threshold. One of the primary challenges for adapting gene set analysis to the microbiome context is the compositional nature of the data. Sequencing technologies constrain the total number of reads, and samples are expected to have the same number of reads instead of DNA content \cite{quinn2019, quinn2018}. However, different samples still yield arbitrarily different total read counts \cite{morton2019, gloor2017}, suggesting the use of normalization methods to allow for proper comparison of feature abundances across samples. However, microbiome data sets do not follow certain assumptions that enable the cross-application of methods from similar fields (such as RNA-seq) \cite{quinn2019, quinn2018}. For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes with constant expression levels across samples. As such, practitioners often rely on total sum normalization to transform count data into relative proportions that sum to one \cite{weiss2017}. Empirical studies have supported this decision, showing increased performance of analysis tasks under this normalization schema \cite{mcknight2019}. Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another. Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1994} are motivated to address this issue.

Standard practice for aggregating variables in microbiome research uses element-wise summations \cite{mclaren2019}. Unfortunately, there are disadvantages to using this approach (referred to as amalgamations in the CoDA literature). First, inter-sample Aitchison distances computed on original parts are not preserved after amalgamation \cite{egozcue2005}. This means that cluster analyses might show different results depending on the level of amalgamation and differs from the those computed from original variables. Second, amalgamations do not allow for comparison between sets of different sizes within the same experimental condition since larger sets will have larger means and variances. Third, considering that each taxa has specific measurement biases \cite{mclaren2019}, an amalgamation based approach would make the bias of the amalgamated variable dependent on the relative abundance of the its constituents. In other words, if taxon $1$ has abundance $A_1$ and bias $B_1$, while taxon $2$ has abundance $A_2$ and bias $B_2$, then the bias of the aggregate variable (for example, a genera) is $(A_1B_1 + A_2B_2)/(A_1 + A_2)$ (see Appendix 1. from McLaren et al. \cite{mclaren2019}). This means that bias invariant approaches (such as analyses of ratios) would no longer be invariant when applied to amalgamated variables as bias now varies across samples. The alternative would be to multiply the proportions rather than to sum them \cite{egozcue2005}.

@@

    \item $\mathbf{X}$: $n$ by $p$ matrix of positive counts for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing. Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing. CBEA does not accept $X$ matrices with zeroes since it invalidates the log-ratio transformation. Users can generate a dense matrix $X$ using a method of choice, however the default mode for CBEA would be add a pseudocount of 1 if zeroes are detected in the matrix. 

@@

    \item \textbf{Calculate finalized CBEA scores with respect to the empirical null}. Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution. P-values can be calculated by subtracting $\mathbf{E}$ from 1. 

@@

\subsubsection*{Estimating the null distribution}
We can assume that the CBEA statistic, similar to other log-ratio based transforms, follows a normal distribution \cite{egozcue2003, aitchison1980}. However, when applying CBEA for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses. Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies. As such, to perform significance testing using CBEA, we also estimated the null distribution from observed raw CBEA variables.

@@

\subsubsection*{Variance inflation due to inter-taxa correlation}  
When taxa within a set are highly correlated, the variance of the sample mean of taxon-wise statistics is inflated. Without loss of generalizability, for a set of taxa with taxa-specific statistics $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:  
\begin{equation} \label{vareq:1}
    Var(\bar{x}) = \frac{1}{m^2}\left(\sum_{i = 1}(\sigma_i^2) + \sum_{i < j}\rho_{ij}\sigma_i\sigma_j\right)
\end{equation}
where $\sigma_i$ is the standard deviation of taxon $i$ while $\rho_{ij}$ is the correlation between $i$ and $j$. The second term of (\ref{vareq:1}) is the correlation dependent variance component, which goes to 0 if there is no correlation. The CBEA statistic follows a similar pattern. Since the geometric mean of a set of variables is equivalent to the exponential of the arithmetic mean of their logarithms, we can re-write CBEA score for a set $k$ with size $K$ as follows:  
\begin{equation}\label{vareq:2}
    M_{i,k} = \sqrt{\frac{K(p - K)}{K + (p - K)}} \left( \overbar{\log{X_{i,j|j \in K}}} - \overbar{\log{X_{i,j|j \notin K}}} \right)   
\end{equation}
where $p$ is the overall number of taxa, $j$ is the index of a taxa and $K$ is the set of indices of taxa in set $k$. The CBEA statistic then looks similar to a t-statistic for difference in means of log-transformed proportions. As such, the pooled variance of cILR is dependent on the variance inflation of both mean components $\overbar{\log{X_{i,j|j \in K}}}$ and $\overbar{\log{X_{i,j|j \notin K}}}$. The result of this variance inflation is inflated type I error since highly correlated sets are also detected as significantly enriched.

@@

\section*{Evaluation} \label{evaluation}
We based our evaluation strategy on gene set testing benchmarking standards set by Geistlinger et al. \cite{geistlinger2021} and utilized the same approaches whenever possible. All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}. All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).

@@

\subsubsection*{Sample-level inference} 
Due to CBEA's self-contained null hypothesis, we can perform inference at the sample level for the enrichment of a set. We evaluated this application by generating one random taxa of different sizes $S \in \{20, 50, 100, 150, 200\}$ across 500 iterations. Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample sets of randomly sampled taxa should not be significantly enriched compared to the remainder background taxa. For this evaluation, we estimated type I error as the fraction of samples that detects our random set as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.  Additionally, this analysis also demonstrate whether CBEA is sensitive to different set sizes.

\subsubsection*{Population-level inference} 
We can perform enrichment testing at the population level by generating corresponding sample level CBEA scores and performing a two-sample test such as Welch's t-test. In order to evaluate CBEA under this context, we generated CBEA scores of sets representing genus-level annotation in above gingival data set \cite{consortium2012, proctor2019} and applied a t-test to test for enrichment (similar to GSVA \cite{hanzelmann2013}) across a randomly generated variable indicating case/control status (repeated 500 times). Type I error is estimated as the fraction of sets per iteration found to be significantly enriched with confidence bands computed from the standard error across all iterations. In addition, we also performed a random set analysis assessing, where we generated 100 sets of different set sizes $S \in \{20, 50, 100, 150, 200\}$ and evaluated the fraction of genera that were found to be differentially abundant across the original labels (supragingival versus subgingival subsite). 95\% confidence intervals were computed using the Agresti-Couli approach \cite{agresti1998}.

\subsection*{Phenotype relevance}
We want to evaluate whether sets found to be significantly enriched by CBEA are relevant to the research question. To perform this assessment, we relied on the gingival data set mentioned prior \cite{consortium2012, proctor2019}. This data set was chosen due to its clear biological interpretation that can serve as the ground truth. Specifically, we expect aerobic microbes to be enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}. Genus-level annotations for microbial metabolism from Beghini et al. \cite{beghini2019} and was obtained from the GitHub repository associated with Calagaro et al. \cite{matteocalgaro2020}. For sample-level inference, we assessed power as the fraction of supragingival samples where aerobic microbes are significantly enriched. For population-level inference, power is the fraction of sets representing genus level taxonomic assignments were significant across subsite labels.

In addition to statistical power, we also assessed phenotype relevance through evaluating whether highly ranked sets based on CBEA scores are more likely to be enriched according to ground truth. This is represented by the area under the receiving operator curve (AUROC/AUC) scores computed on CBEA scores against true labels (similar approach was used to evaluate VAM \cite{frost2020}). DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate.

@@

\begin{enumerate}
\item Whole genome sequencing of stool samples from inflammatory bowel disease (IBD) patients in the MetaHIT consortium \cite{nielsen2014}. This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease). We processed this data by removing all samples with less than 1,000 total read counts as well as any OTU who was present (at least 1 count) in 10\% of samples or less. Prior to model fitting, we back-transformed relative abundances into count data (to align the format with our 16S rRNA gene sequencing data set) using provided total number of reads aligned to MetaPhlan marker genes (per sample).

@@

\subsection*{Comparison Methods} 
We benchmarked the statistical properties of CBEA against existing baseline approaches. For sample-level inference analyses, utilized the Wilcoxon rank-sum test, which non-parametrically tests the difference in mean counts between taxa from a pre-defined set and its remainder similar to CBEA. For assessments at the population level, we compared CBEA against performing a standard test for differential abundance with set-level features generated via element-wise summations instead. We chose DESeq2 \cite{love2014} and corncob \cite{martin2020} since they represent both methods extrapolated from RNA-seq \cite{mcmurdie2014} and those developed specifically for microbiome data.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width = \textwidth]{figures/data_ss_fdr_new.png}
    \caption{Random taxa set analyses for inference at the sample level of CBEA under different parametric assumptions compared against a Wilcoxon rank-sum test. Type I error (\emph{y}-axis) was evaluated by generating random sets of different sizes (\emph{x}-axis) (500 replications per size) and computing the fraction of samples where the set was found to be significantly enriched at $\alpha = 0.05$. Error bars represent the mean type I error $\pm$ sample standard error computed across 500 replications of the experiment.} 
    \label{fig:2}
\end{figure}

@@

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width = \textwidth]{figures/data_diffab_fdr.eps}
    \caption{Random sample label (\textbf{A}) and random set (\textbf{B}) analyses for population level inference. (\textbf{A}) Type I error (\emph{x}-axis) was estimated as the overall fraction of sets found to be enriched $\alpha = 0.05$ using randomly generated sample labels (500 permutations).  Error bars represent the mean type I error $\pm$ sample standard error. (\textbf{B}) Proportion of significant sets (\emph{y}-axis) using 100 randomly generated sets of different set sizes (\emph{x}-axis). Confidence intervals computed using Agresti-Couli method for binomial proportions.}  
    \label{fig:3}
\end{figure}

@@

We also assessed the impact of set-size on the inference procedure by testing for enrichment using the original sample labels but with randomly sampled sets of different sizes. Overall we observed very similar values across CBEA as well as corncob and DESeq2, suggesting that no individual method is systematically identifying too many significant sets. Additionally, similar to analogous analyses at the sample level, no approach were significantly sensitive to set-size.

@@

We estimated statistical power using this data set as the fraction of supragingival samples where the set representing aerobic microbes were significantly enriched. We observed that adjusted CBEA approaches demonstrate much lower power compared to the Wilcoxon rank-sum test and unadjusted variants. This is surprising given the fact that in statistical significance analyses, the adjusted CBEA approach provides inflated type I error, especially if the normal distribution assumption was chosen. Here, we did not see any differences in performance between distribution choices.

@@

Unfortunately, evaluating power and score rankings using real data does not allow for observing performance trends when effect sizes, sparsity, and inter-taxa correlation are varied. In equivalent simulation analyses, we observed the opposite effect, where

@@

\subsection*{Disease Prediction}   
Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models. In this section we assessed the predictive performance of a standard baseline random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating CBEA, ssGSEA, and GSVA). Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width = \linewidth]{figures/data_prediction_plot.png}
    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles. Data sets used span both 16S rRNA gene sequencing (Gevers et al. \cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al. \cite{nielsen2014}). CBEA performs better than GSVA and ssGSEA but not as well as CLR, with the exception of the whole genome sequencing data set.}
    \label{fig:6}
\end{figure}

Fig~\ref{fig:6} illustrates results from using CBEA, ssGSEA, GSVA, or CLR transformed variables as inputs to our model, with AUROC as the performance criteria. Results also replicated that of the simulations (Supplementary materials), where across both data sets CBEA and CLR methods provide much better performance than both GSVA or ssGSEA. Interestingly, the CBEA approach performed better than CLR in the whole genome data set but did not perform as well in the 16S rRNA gene sequencing data set. However, these results indicate that CBEA generated scores are informative, providing competitive performance when acting as inputs to disease predictive models. Most importantly, performance values are consistent across both simulated and real data sets.

These results demonstrate for given set-based features, CBEA can generate scores can could be informative in disease prediction tasks. Simulation results indicate that CBEA methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach. Interestingly, however, CBEA methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.

@@

For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}) while also demonstrating respectable power (Fig~\ref{fig:4}). This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), where the mixture normal displayed superior fit to the null distribution. Unfortunately, other variants of CBEA did not demonstrate good type I error control. Interestingly, while the adjusted methods showed inflated type I error and low power in real data evaluations (Fig~\ref{fig:2}), they performed well in simulation studies (\nameref{S1_Fig}, \nameref{S3_Fig}).

>>>>

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing. However, microbiome taxonomic data is sparse and high dimensional which prevents the application of standard statistical techniques. One approach to address this problem is to perform analyses at the level of taxon sets. Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability. Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given its unique features compared to other 'omics data types. We developed a new tool to generate taxon set enrichment scores at the sample level through a novel log-ratio formulation based on the competitive null hypothesis. Our scores can be used for statistical inference at both the sample and population levels, as well as inputs to other downstream analyses such as prediction models. We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.

@@

In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate taxa to higher Linnean classification levels such as genus, family, or phylum. However, despite this interest, there is limited research done to extend existing set-based methods to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used and p-value threshold. One of the primary challenges for adapting gene set analysis to the microbiome context is the compositional nature of the data. Sequencing technologies constrain the total number of reads, and samples are expected to have the same number of reads instead of DNA content \cite{quinn2019, quinn2018}. However, different samples still yield arbitrarily different total read counts \cite{morton2019, gloor2017}, suggesting the use of normalization methods to allow for proper comparison of feature abundances across samples. However, microbiome data sets do not follow certain assumptions that enable the cross-application of methods from similar fields (such as RNA-seq) \cite{quinn2019, quinn2018}. For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes with constant expression levels across samples. As such, practitioners often rely on total sum normalization to transform count data into relative proportions that sum to one \cite{weiss2017}. Some studies have provided emprical performance vealuations supporting this normalization schema \cite{mcknight2019}. Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another. Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1994} are motivated to address this issue.

Unfortunately, the standard practice for aggregating variables using element-wise summations (referred to as amalgamations in the CoDA literature), does not adequately address the compositional issue \cite{mclaren2019}. First, inter-sample Aitchison distances computed on original parts are not preserved after amalgamation \cite{egozcue2005}. This means that cluster analyses might show different results depending on the level of amalgamation and differs from the those computed from original variables. Second, amalgamations do not allow for comparison between sets of different sizes within the same experimental condition since larger sets will have larger means and variances. Third, considering that each taxa has specific measurement biases \cite{mclaren2019}, an amalgamation based approach would make the bias of the amalgamated variable dependent on the relative abundance of the its constituents. In other words, if taxon $1$ has abundance $A_1$ and bias $B_1$, while taxon $2$ has abundance $A_2$ and bias $B_2$, then the bias of the aggregate variable (for example, a genera) is $(A_1B_1 + A_2B_2)/(A_1 + A_2)$ (see Appendix 1. from McLaren et al. \cite{mclaren2019}). This means that bias invariant approaches (such as analyses of ratios) would no longer be invariant when applied to amalgamated variables as bias now varies across samples. The alternative would be to multiply the proportions rather than to sum them \cite{egozcue2005}.

@@

    \item $\mathbf{X}$: $n$ by $p$ matrix of positive propotions for $p$ taxa and $n$ samples measured through either targeted sequencing (such as of the 16S rRNA gene) or whole genome shotgun sequencing. Usually $\mathbf{X}$ is generated from standard taxonomic profiling pipelines such as DADA2 \cite{callahan2016} for 16S rRNA sequencing, or MetaPhlAn2 \cite{truong2015} for whole genome shotgun sequencing. CBEA does not accept $X$ matrices with zeroes since it invalidates the log-ratio transformation. Users can generate a dense matrix $X$ using a method of choice, however the default mode for CBEA would be add a pseudocount of $10^{-5}$ if zeroes are detected in the matrix. 

@@

    \item \textbf{Calculate finalized CBEA scores with respect to the empirical null}. Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution. Raw p-values can be calculated by subtracting $\mathbf{E}$ from 1. 

@@

\subsubsection*{Estimating the null distribution}
We can assume that the CBEA statistic, similar to other log-ratio based transforms, follows a normal distribution \cite{egozcue2003, aitchison1980}. However, when applying CBEA for hypothesis testing at the sample level, it is expected that the researcher would be testing a large number of hypotheses. Under the assumption that the number of truly significant hypotheses is low, Efron \cite{efron2004} showed that estimating the null distribution of the test statistic directly (termed the empirical null distribution) is much more preferable than using the theoretical null due to unobserved confounding effects inherently part of observational studies. As such, to perform significance testing using CBEA, we also estimated the null distribution from observed raw CBEA variables.

@@

\subsubsection*{Variance inflation due to inter-taxa correlation}  
When taxa within a set are highly correlated, the variance of the sample mean of taxon-wise statistics is inflated. Without loss of generalizability, for a set of taxa with taxa-specific statistics $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:  
\begin{equation} \label{vareq:1}
    Var(\bar{x}) = \frac{1}{m^2}\left(\sum_{i = 1}(\sigma_i^2) + \sum_{i < j}\rho_{ij}\sigma_i\sigma_j\right)
\end{equation}
where $\sigma_i$ is the standard deviation of taxon $i$ while $\rho_{ij}$ is the correlation between $i$ and $j$. The second term of (\ref{vareq:1}) is the correlation dependent variance component, which goes to 0 if there is no correlation. The CBEA statistic follows a similar pattern. Since the geometric mean of a set of variables is equivalent to the exponential of the arithmetic mean of their logarithms, we can re-write CBEA score for a set $k$ with size $K$ as follows:  
\begin{equation}\label{vareq:2}
    M_{i,k} = \sqrt{\frac{K(p - K)}{K + (p - K)}} \left( \overbar{\log{X_{i,j|j \in K}}} - \overbar{\log{X_{i,j|j \notin K}}} \right)   
\end{equation}
where $p$ is the overall number of taxa, $j$ is the index of a taxa and $K$ is the set of indices of taxa in set $k$. The CBEA statistic then looks similar to a t-statistic for difference in means of log-transformed proportions. As such, the pooled variance of CBEA is dependent on the variance inflation of both mean components $\overbar{\log{X_{i,j|j \in K}}}$ and $\overbar{\log{X_{i,j|j \notin K}}}$. The result of this variance inflation is inflated type I error since highly correlated sets are also detected as significantly enriched.

@@

\section*{Evaluation} \label{evaluation}
We based our evaluation strategy on gene set testing benchmarking standards set by Geistlinger et al. \cite{geistlinger2021} and utilized the same approaches whenever possible. All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}. All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}). Additional packages used to support this analysis includes: \emph{tidyverse} \cite{wickham2019}, \emph{pROC} \cite{robin2011}, \emph{phyloseq} \cite{mcmurdie2014}, \emph{mia} \cite{ernst2021}, \emph{targets} \cite{landau2021}.

@@

\subsubsection*{Sample-level inference} 
Due to CBEA's self-contained null hypothesis, we can perform inference at the sample level for the enrichment of a set. We evaluated this application by generating one random taxa of different sizes $S \in \{20, 50, 100, 150, 200\}$ across 500 iterations. Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample sets of randomly drawn taxa should not be significantly enriched compared to the remainder background taxa. For this evaluation, we estimated type I error as the fraction of samples that detects our random set as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.  Additionally, this analysis also demonstrate whether CBEA is sensitive to different set sizes.

\subsubsection*{Population-level inference} 
We can perform enrichment testing at the population level by generating corresponding sample level CBEA scores and performing a two-sample test such as Welch's t-test. In order to evaluate CBEA under this context, we generated CBEA scores of sets representing genus-level annotation in above gingival data set \cite{consortium2012, proctor2019} and applied a t-test to test for enrichment (similar to GSVA \cite{hanzelmann2013}) across a randomly generated variable indicating case/control status (repeated 500 times). Type I error is estimated as the fraction of sets per iteration found to be significantly enriched with confidence bands computed from the standard error across all iterations. In addition, we also performed a random set analysis assessment, where we generated 100 sets of different set sizes $S \in \{20, 50, 100, 150, 200\}$ and evaluated the fraction of genera that were found to be differentially abundant across the original labels (supragingival versus subgingival subsite). 95\% confidence intervals were computed using the Agresti-Couli approach \cite{agresti1998}.

\subsection*{Phenotype relevance}
We want to evaluate whether sets found to be significantly enriched by CBEA are relevant to the research question. To perform this assessment, we relied on the gingival data set mentioned prior \cite{consortium2012, proctor2019}. This data set was chosen due to its clear biological interpretation that can serve as the ground truth. Specifically, we expect aerobic microbes to be enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}. Genus-level annotations for microbial metabolism from Beghini et al. \cite{beghini2019} and was obtained from the GitHub repository associated with Calagaro et al. \cite{matteocalgaro2020}. For sample-level inference, we assessed power as the fraction of supragingival samples where aerobic microbes are significantly enriched. For population-level inference, power is the fraction of sets representing genus level taxonomic assignments that were significant across subsite labels.

In addition to statistical power, we also assessed phenotype relevance through evaluating whether highly ranked sets based on CBEA scores are more likely to be enriched according to ground truth. This is represented by the area under the receiving operator curve (AUROC/AUC) scores computed on CBEA scores against true labels (similar approach was used to evaluate VAM \cite{frost2020}). DeLong 95\% confidence intervals for AUROC \cite{delong1988} were obtained for each estimate.

@@

\begin{enumerate}
\item Whole genome sequencing of stool samples from inflammatory bowel disease (IBD) patients in the MetaHIT consortium \cite{nielsen2014}. This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease). We processed this data by removing all samples with less than 1,000 total read counts as well as any OTU who was present (with non-zero proportions) in 10\% of samples or less. Prior to model fitting, we back-transformed relative abundances into count data (to align the format with our 16S rRNA gene sequencing data set) using provided total number of reads aligned to MetaPhlan marker genes (per sample).

@@

\subsection*{Comparison Methods} 
We benchmarked the statistical properties of CBEA against existing baseline approaches. For sample-level inference analyses, utilized the Wilcoxon rank-sum test, which non-parametrically tests the difference in mean counts between taxa from a pre-defined set and its remainder similar to CBEA. For assessments at the population level, we compared CBEA against performing a standard test for differential abundance with set-level features generated via element-wise summations instead. We chose DESeq2 \cite{love2014} and corncob \cite{martin2020} because they represent both methods extrapolated from RNA-seq \cite{mcmurdie2014} and those developed specifically for microbiome data.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width = \textwidth]{figures/data_ss_fdr_new.png}
    \caption{Random taxa set analyses for inference at the sample level of CBEA under different parametric assumptions compared against a Wilcoxon rank-sum test. Type I error (\emph{y}-axis) was evaluated by generating random sets of different sizes (\emph{x}-axis) (500 replications per size) and computing the fraction of samples where the set was found to be significantly enriched at $\alpha = 0.05$. Error bars represent the mean type I error $\pm$ sample standard error computed across 500 replications of the experiment. Only the unadjusted CBEA with the mixture normal distribution and the Wilcoxon rank sum test were able to control for type I error at 0.05. All approaches are invariant to set sizes.} 
    \label{fig:2}
\end{figure}

@@

Interestingly, simulation results (\nameref{S1_Fig}) showed an opposite pattern. Adjusted approaches were good at controlling for type I error, especially under the low inter-taxa correlation values within the set (similar to generating random sets where the natural correlation structure is disrupted). In these simulations, unadjusted approaches and the Wilcoxon rank sum test had significant type I error inflation with increasing correlation. All approaches seems to be invariant to the level of data sparsity.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width = \textwidth]{figures/data_diffab_fdr.eps}
    \caption{Random sample label (\textbf{A}) and random set (\textbf{B}) analyses for population level inference. (\textbf{A}) Type I error (\emph{x}-axis) was estimated as the overall fraction of sets found to be enriched $\alpha = 0.05$ using randomly generated sample labels (500 permutations).  Error bars represent the mean type I error $\pm$ sample standard error. (\textbf{B}) Proportion of significant sets (\emph{y}-axis) using 100 randomly generated sets of different set sizes (\emph{x}-axis). Confidence intervals computed using Agresti-Couli method for binomial proportions. For sample label permutation (\textbf{A}), all CBEA approaches were able to control for type I error but not for corncob and DESeq2. For random set analyses (\textbf{B}), all approaches demonstrate similar rate of accepting significant sets and were invariant to overall set size.}  
    \label{fig:3}
\end{figure}

@@

We also assessed the impact of set-size on the inference procedure by testing for enrichment using the original sample labels but with randomly sampled sets of different sizes (Fig~\ref{fig:3}B). Overall we observed very similar values across CBEA as well as corncob and DESeq2, suggesting that no individual method is systematically identifying too many significant sets. Additionally, similar to analogous analyses at the sample level, no approach were significantly sensitive to changes in set sizes.

@@

We estimated statistical power using this data set as the fraction of supragingival samples where the set representing aerobic microbes were significantly enriched. We observed that adjusted CBEA approaches demonstrate much lower power compared to the Wilcoxon rank-sum test and unadjusted variants. This is surprising given the fact that in statistical significance analyses, the adjusted CBEA approach provides inflated type I error, especially if the normal distribution assumption was chosen, which indicates a mismatch in estimating the null distribution since a high type I error did not result in increased power.

@@

The above results were replicated in simulation studies where we observed that adjusted approaches was very conservative and demonstrated significantly lower power (\nameref{S3_Fig}) with increasing correlation even at the highest evaluated effect sizes. When assessing score rankings, the performance of CBEA was closer to ssGSEA and GSVA compared to real data evaluations, however all single sample approaches were much better than using the W statistic from the Wilcoxon Rank Sum test.

@@

\subsection*{Disease Prediction}   
Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:4}), we want to assess whether they can also act as useful inputs to predictive models. In this section we assessed the predictive performance of a standard baseline random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (CBEA, ssGSEA, and GSVA). Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width = \linewidth]{figures/data_pred.png}
    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on predicting patients with inflammatory bowel disease versus controls using genus level taxonomic profiles. Data sets used span both 16S rRNA gene sequencing (Gevers et al. \cite{gevers2014}) and whole-genome shotgun sequencing (Nielsen et al. \cite{nielsen2014}). CBEA performs better than GSVA and ssGSEA but not as well as CLR, with the exception of the whole genome sequencing data set.}
    \label{fig:6}
\end{figure}

Fig~\ref{fig:6} illustrates the performance of our model with AUROC as the evaluation criteria. In the 16S rRNA data set, the best performing CBEA variant (CDF values computed from an unadjusted mixture normal distribution) outperforms both GSVA and ssGSEA but not the standard CLR approach. Interestingly, in the whole genome sequencing data set, CBEA outperforms CLR, but were more similar in performance to GSVA. However, due to large confidence intervals, no method were significantly out-performing other approaches evaluated. As such, these results indicated that CBEA generated scores are can be informative, providing competitive performance when acting as inputs to disease predictive models. Simulation studies (\nameref{S5_Fig}) showed similar results, however CBEA was more consistently underperforming compared to CLR across all scenarios. Interestingly, the performance gap decreases with increasing sparsity levels.

@@

For single-sample analyses, we demonstrated that the CBEA approach (unadjusted with mixture normal parametric assumption) was able to control for type I error at the nominal level of 0.05 under the global null (Fig~\ref{fig:2}) while also demonstrating respectable power (Fig~\ref{fig:4}). This performance is consistent across different set sizes as well as our prior distributional fit analyses (Fig~\ref{fig:1}), where the mixture normal displayed superior fit to the null distribution. Unfortunately, other variants of CBEA demonstrated neither good type I error control nor power. Interestingly, while the adjusted methods showed poor performance in real data evaluations (Fig~\ref{fig:2}), in simulation studies (\nameref{S1_Fig}, \nameref{S3_Fig}) these approaches were able to control for type I error well with the trade-off of much lower power.
